---
title: "Getting started with ggsegverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggsegverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Neuroimaging analysis often ends with a table of regions and p-values.
The hard part is turning that table into a figure your audience can actually read.
ggsegverse exists to close that gap — one `library()` call loads everything you need to go from statistics to publication-ready brain figures in R.

```{r}
#| label: load
library(ggsegverse)
```

## What you get

ggsegverse attaches five packages, each handling a different piece of the visualization pipeline:

- **ggseg** — a ggplot2 geom for 2D brain region plots. If you can build a ggplot, you can build a brain figure.
- **ggseg3d** — interactive 3D brain widgets built on Three.js, with `snapshot_brain()` for static images.
- **ggseg.formats** — the `brain_atlas` and `ggseg_atlas` data structures that both ggseg and ggseg3d rely on, plus bundled atlases (Desikan-Killiany, FreeSurfer aseg, TRACULA, SUIT cerebellum) and FreeSurfer statistics readers.
- **ggseg.meshes** — additional cortical surfaces (pial, white, midthickness, and more) and cerebellar meshes for 3D rendering.
- **ggplot2** — the grammar of graphics that ggseg builds on, so `theme()`, `scale_fill_*()`, `labs()` and friends are ready to use.

You do not need to remember which function lives where.
Loading ggsegverse makes them all available.

On top of attaching those packages, ggsegverse manages the ecosystem for you.
`ggseg_atlas_repos()` and `installed_ggseg_atlases()` tell you which atlas packages exist and which ones you already have, `install_ggseg_atlas()` installs them, and `ggsegverse_sitrep()`, `ggsegverse_deps()` and `ggsegverse_conflicts()` tell you what state your installation is in.
Those are covered below.

ggsegverse is about plotting.
If you want to build your own atlas from a parcellation, that is the job of [ggseg.extra](https://ggsegverse.github.io/ggseg.extra/), which you install separately.

## A first plot

The ecosystem ships four atlases out of the box: `dk()` (Desikan-Killiany cortical parcellation), `aseg()` (FreeSurfer subcortical segmentation), `tracula()` (white matter tracts), and `suit()` (cerebellar lobules).
These are accessed as function calls that return atlas objects.

A cortical surface plot uses `geom_brain()`:

```{r}
#| label: dk-basic
#| fig-alt: >
#|   2D cortical brain plot showing the Desikan-Killiany atlas regions
#|   colored by label
#| fig-cap: "The Desikan-Killiany atlas rendered with ggseg's default settings."
ggplot() +
  geom_brain(atlas = dk(), show.legend = FALSE)
```

The subcortical segmentation works the same way:

```{r}
#| label: aseg-basic
#| fig-alt: "2D axial view of subcortical brain structures from the aseg atlas"
#| fig-cap: "FreeSurfer's automatic subcortical segmentation atlas."
ggplot() +
  geom_brain(atlas = aseg(), show.legend = FALSE)
```

## Mapping your own data

The real value is mapping _your_ data onto brain regions.
Say you have cortical thickness values for a subset of regions:

```{r}
#| label: join-data
#| fig-alt: >
#|   Cortical plot with four regions colored by thickness values on a
#|   magma gradient
#| fig-cap: "Cortical thickness mapped onto Desikan-Killiany regions."
my_data <- data.frame(
  region = c(
    "precentral",
    "postcentral",
    "superiorfrontal",
    "inferiorparietal"
  ),
  thickness = c(2.6, 2.1, 2.8, 2.4)
)

ggplot() +
  geom_brain(
    data = my_data,
    atlas = dk(),
    mapping = aes(fill = thickness),
    show.legend = TRUE
  ) +
  scale_fill_viridis_c(option = "A") +
  labs(fill = "Thickness (mm)") +
  theme_brain()
```

`geom_brain()` joins your data to the atlas by the `region` column.
Regions without data stay grey.

## Interactive 3D views

For presentations or exploratory work, `ggseg3d()` gives you a rotatable Three.js widget:

```{r}
#| label: ggseg3d-basic
#| eval: false
ggseg3d(atlas = dk())
```

Add a translucent cortical surface for anatomical context around subcortical structures:

```{r}
#| label: glassbrain
#| eval: false
ggseg3d(atlas = aseg()) |>
  add_glassbrain(opacity = 0.15) |>
  pan_camera("right lateral")
```

## Finding and installing atlases

The bundled atlases cover the basics, but the ggsegverse [r-universe](https://ggsegverse.r-universe.dev) hosts many more — Yeo networks, Schaefer parcellations, Glasser HCP-MMP, Harvard-Oxford, and others.

Browse what is available:

```{r}
#| label: browse-atlases
#| eval: false
ggseg_atlas_repos()
```

This queries the ggsegverse r-universe and returns a tibble of atlas packages with their versions and descriptions.
Core packages and tools hosted there, like ggseg.extra, are left out.
You can narrow the list with a pattern:

```{r}
#| label: filter-atlases
#| eval: false
ggseg_atlas_repos("yeo")
```

### Installing an atlas

Once you have found what you need, `install_ggseg_atlas()` handles the r-universe repository setup for you:

```{r}
#| label: install-atlas
#| eval: false
install_ggseg_atlas("ggsegYeo2011")
```

After installation, the atlas is available like any other R package:

```{r}
#| label: use-installed-atlas
#| eval: false
library(ggsegYeo2011)
ggplot() +
  geom_brain(atlas = yeo7(), show.legend = FALSE)
```

### Installing everything

If disk space is not a concern and you want the full collection:

```{r}
#| label: install-all
#| eval: false
install_ggseg_atlas_all()
```

This is a large download, so it asks you to confirm first.
In a script or a non-interactive session there is nobody to ask, so you have to opt in explicitly:

```{r}
#| label: install-all-scripted
#| eval: false
install_ggseg_atlas_all(ask = FALSE)
```

### Checking what you have

`installed_ggseg_atlases()` lists the atlas packages you have installed, with your local version next to the latest one on the r-universe:

```{r}
#| label: installed-atlases
#| eval: false
installed_ggseg_atlases()
```

## Keeping things current

`ggsegverse_sitrep()` gives you a quick diagnostic of the core packages — what is installed, what version you are running, and whether anything is out of date:

```{r}
#| label: sitrep
#| eval: false
ggsegverse_sitrep()
```

If updates are available, `ggsegverse_update()` prints the exact `pak::pak()` call to run.
It does not install anything for you:

```{r}
#| label: update
#| eval: false
ggsegverse_update()
```

## Name conflicts

When ggsegverse is attached, it lists any function that a core package masks, or that masks one of its functions.
Re-exports of the same function across ecosystem packages, like `dk()`, are not counted as conflicts.
You can re-run the check at any time:

```{r}
#| label: conflicts
ggsegverse_conflicts()
```

If you prefer ambiguous calls to fail loudly, the [conflicted](https://conflicted.r-lib.org/) package turns every conflict into an error until you choose a winner.

## Where to go next

- [ggseg documentation](https://ggsegverse.github.io/ggseg/) — full reference for 2D plotting, custom themes, and data joining
- [ggseg3d documentation](https://ggsegverse.github.io/ggseg3d/) — 3D widget customization, camera control, and static snapshots
- [ggseg.formats documentation](https://ggsegverse.github.io/ggseg.formats/) — building and converting atlas objects
- [ggseg.extra documentation](https://ggsegverse.github.io/ggseg.extra/) — building your own atlases (not attached by ggsegverse; install it separately when you need it)
